Examining the case behind Acer Platanoide's plight on Toronto's Street Trees 

### 2024-12-03 paper/paper.qmd
Updated paper, and reference file. Included re-render

Urban spaces can be void of vast expanses of nature, whether it be grass fields, blooming flowers, or teaming and diverse natural wildlife. One simple way to tackle this is to incorporate nature within the urban landscape, such as with trees. Trees are a vital plant in the world, they take in carbon dioxide and provide oxygen, a key component in the air we breath. Urban areas are known for the excess in carbon dioxide, which only makes it seem natural that trees can live and perhaps thrive in such environments. One way cities have incorporated trees into the city are roadside trees. Trees line many of the cities highways, arterials, collectors, and small roads. Lines of trees lined alongside highways can act as a natural sound barrier by reflecting, scattering and absorbing nearby noises (@aais). Tree-lined streets can also alter the way wind affects pedestrians below them, creating a more hospitable environment to promote active transportation such as walking or biking (@beptree). It should be no surprise then, that many cities around the world, including Toronto, have incorporated trees into their landscape, in busy downtown centres, to sleepy suburbs, and everywhere in between.

Acer platanoides is a well known tree used in urban environments, especially after the demise of elm trees in the 1970s (@urbanforest). These trees are know for ‘a vigorous early growth rate, desirable form and size, the capacity to withstand many urban impacts (e.g. pavement, moderate levels of pollution, dusts, and dry soils) and the abilities to transplant well, grow on a wide variety of soils, and withstand ice and snow damage better than other maples’ (@norwaymaple). Being planted in cities, it's seeds can travel by wind across to more wooded suburbs, where the seeds can tolerate shade for extended periods of time (@urbanforest). However, as it's common name suggests, it is not a native species to Canada, instead brought here by European explorers and settlers. While not a very invasive species, taking over just 9% of Southern Ontario's forests, it does so by harming other native plants and trees (@urbanforests). More attention should be cast on the spread of this tree, as it has the potential to devastate Toronto's and Southern Ontario's forests in the years to come.

@@

All collection and analysis of data was done using R, a free programming language designed for data scientists (@citeR). Additionally, tidycerse (@citeTidyverse), dplyr (@citeDplyr), and janitor (@citeJanitor) have been used as packages to read, clean, and present data. Finally OpenData Toronto is an online database containing thousands of datasets pertaining to the City of Toronto, which is where the data for this paper was initially accessed (@citeOpendatatoronto).

Rohan Alexander's book, *Telling Stories with Data* was referenced for troubleshooting and general ideas for data analysis (@tellingstories)

The data comes from Urban Forestry, an organization dedicated to Toronto's urban forests. They work towards planting more trees, and protecting existing trees from damage due to individuals, private entities, and public entities (@urbanforestry). This particular dataset is called Street Tree Data, and focuses on city-owned trees located on roads. This information would be most likely used by city planners and road maintenance, ensuring snow plows and street dusters don't impede on any trees that line the roads. It is important to note that there may be some privately owned trees listed among the municipal trees, as they may be of interest with regards to road maintenance or other city services.

Street Tree Data contains around 32000 observations, each observation being an individual tree. Each observation has a general id for data analysis, as well as an structure id, telling us if the tree is part of an existing structure or building. Location is covered in 4 ways: first the address number (address) and streetname (streetname) of nearest postal delivery address, next is the street name and the nearest 2 cross streets (crosstreet1 and crossstreet2), third being the ward number (ward), and finally geometry containain a tuple of general latitude and longitude. Street Tree contains both the species name (botanical_name) and the common English name common_name), however we have decided to use only the species name to avoid confusion between similar sounding trees. Common English names will be provided for species of key interest. There is also measurement called dbh_trunk. This is the standard method forestry experts use to measure the diameter of a tree. Depending on where the measurement is taking placce, the breast height changes. For example, in Canada it is 1.3 meters, while the US uses 4.5 feet. This methodology is comprised of old conventions used to measure trees used throughout the world (@dbh). The result is a complicated mess where breast_height is a vague term that changes by location. Additionally, the convention breaks down when the tree is anything but vertically upright and straight. As a result, the DBH is more of an estimate rather than a sturdy measurement.

@@

## `ward`
Ward is one of several location varaibles included in the Street Trees dataset. It lists the ward location of each observation (tree) in the city. Toronto is broken up into municipal wards, who is represented in City Hall by a councillor. Wards vary in size and population, which may have an impact in the spread of Acer platanoides. For example, one of the largest Wards is Etobicoke North, which forms the north-western boundary of Toronto. Given its size and distance from downtown Toronto, it has ample room to host large swaths of forests, and have its roads lined with trees, increasing the possibility of Acer platanoides finding its way here. Toronto-Centre is the smallest ward by far, and being situated in downtown Toronto, is more dense with human infrastructure. There is less space for Acer platanoides to grow here, but the few green spots here could be suceptible to Acer platanoides.

@@

@@

Next steps would be augmenting this data with other data from Toronto's roads, or other trees, to see if any comparisions or correlations can be drawn there. Does vehicle traffic impact the trees lining a certain road, and if so, how long would it take for the effects to be noticeable? Does the distribution of street trees tell us anything about income disparity, fire risk, or quality of life within Toronto? This report can be used as a launching pad for further investigations, using key takeways from this report as drivers for further questions.

With newer report, pressure will be put upon Urban Forestry to better curtail the data they have collected thus far. It is imperative that their hard work be recognized and any feedback on the data they have collected be sent to them as well, ensuring better datasets for analysts, leading to more informative conlusions for all.

@@

@@

While cleaning that dataset, some inconsisten values were discovered. A particular species of tree, Salix x sepulcralis, had it's name misspelled in the dataset. This may have come from using special characters that could not be saved onto the dataset, resulting in a garbled name. An additional step was taken to restore the correct botanical name, using the common name as reference. Similarly, one observation did not have a ward location allocated. This was remedied by reading the globval coordinates from geometry, a varaible in the raw data file that shares the latitude and longtitude of the tree individual. Using these coordinates, and the address number, the general location for this tree was determined, along with its ward location. The ward entry for this observataion was updated accordingly.

>>>>

Urban spaces can be void of vast expanses of nature, whether it be grass fields, blooming flowers, or teaming and diverse natural wildlife. One simple way to tackle this is to incorporate nature within the urban landscape, such as with trees. Trees are a vital plant in the world, they take in carbon dioxide and provide oxygen, a key component in the air we breath. Urban areas are known for the excess in carbon dioxide, which only makes it seem natural that trees can live and perhaps thrive in such environments. One way cities have incorporated trees into the city are roadside trees. Trees line many of the cities highways, arterial roads, collector roads, and small roads. Lines of trees lined alongside highways can act as a natural sound barrier by reflecting, scattering and absorbing nearby noises (@aais). Tree-lined streets can also alter the way wind affects pedestrians below them, creating a more hospitable environment to promote active transportation such as walking or biking [@beptree[. It should be no surprise then, that many cities around the world, including Toronto, have incorporated trees into their landscape, in busy downtown centres, to sleepy suburbs, and everywhere in between.

Acer platanoides is a well known tree used in urban environments, especially after the demise of elm trees in the 1970s [@urbanforest]. These trees are know for ‘a vigorous early growth rate, desirable form and size, the capacity to withstand many urban impacts (e.g. pavement, moderate levels of pollution, dusts, and dry soils) and the abilities to transplant well, grow on a wide variety of soils, and withstand ice and snow damage better than other maples’ [@norwaymaple]. Being planted in cities, it's seeds can travel by wind across to more wooded suburbs, where the seeds can tolerate shade for extended periods of time [@urbanforest]. However, as it's common name suggests, it is not a native species to Canada, instead brought here by European explorers and settlers. While not a very invasive species, taking over just 9% of Southern Ontario's forests, it does so by harming other native plants and trees [@urbanforests]. More attention should be cast on the spread of this tree, as it has the potential to devastate Toronto's and Southern Ontario's forests in the years to come.

@@

All collection and analysis of data was done using R, a free programming language designed for data scientists [@citeR]. Additionally, tidycerse [@citeTidyverse], dplyr [@citeDplyr], and janitor [@citeJanitor] have been used as packages to read, clean, and present data. Finally OpenData Toronto is an online database containing thousands of datasets pertaining to the City of Toronto, which is where the data for this paper was initially accessed [@citeOpendatatoronto].

Rohan Alexander's book, *Telling Stories with Data* was referenced for troubleshooting and general ideas for data analysis [@tellingstories]

The data comes from Urban Forestry, an organization dedicated to Toronto's urban forests. They work towards planting more trees, and protecting existing trees from damage due to individuals, private entities, and public entities [@urbanforestry]. This particular dataset is called Street Tree Data, and focuses on city-owned trees located on roads. This information would be most likely used by city planners and road maintenance, ensuring snow plows and street dusters don't impede on any trees that line the roads. It is important to note that there may be some privately owned trees listed among the municipal trees, as they may be of interest with regards to road maintenance or other city services.

Street Tree Data contains around 32000 observations, each observation being an individual tree. Each observation has a general id for data analysis, as well as an structure id, telling us if the tree is part of an existing structure or building. Location is covered in 4 ways: first the address number (address) and streetname (streetname) of nearest postal delivery address, next is the street name and the nearest 2 cross streets (crosstreet1 and crossstreet2), third being the ward number (ward), and finally geometry contains a tuple of general latitude and longitude. Street Tree contains both the species name (botanical_name) and the common English name common_name), however we have decided to use only the species name to avoid confusion between similar sounding trees. Common English names will be provided for species of key interest. There is also measurement called dbh_trunk. This is the standard method forestry experts use to measure the diameter of a tree. Depending on where the measurement is taking place, the breast height changes. For example, in Canada it is 1.3 meters, while the US uses 4.5 feet. This methodology is comprised of old conventions used to measure trees used throughout the world [@dbh]. The result is a complicated mess where breast_height is a vague term that changes by location. Additionally, the convention breaks down when the tree is anything but vertically upright and straight. As a result, the DBH is more of an estimate rather than a sturdy measurement.

@@

## `ward`
Ward is one of several location varaibles included in the Street Trees dataset. It lists the ward location of each observation (tree) in the city. Toronto is broken up into municipal wards, who is represented in City Hall by a Councillor. Wards vary in size and population, which may have an impact in the spread of Acer platanoides. For example, one of the largest Wards is Etobicoke North, which forms the north-western boundary of Toronto. Given its size and distance from downtown Toronto, it has ample room to host large swaths of forests, and have its roads lined with trees, increasing the possibility of Acer platanoides finding its way here. Toronto-Centre is the smallest ward by far, and being situated in downtown Toronto, is more dense with human infrastructure. There is less space for Acer platanoides to grow here, but the few green spots here could be susceptible to Acer platanoides.

@@

# Measurement
Urban Forestry Management oversees trees within the City of Toronto. Data used in Street Trees comes from a program called Area Street Tree Maintenance (ASTM) [@StreetTreeMaintenance]. ASTM uses a form of cluster sampling to collect data and preform maintenance on street trees. Forestry Management gets information on a general area, which we can call a cluster. Within that cluster, each street tree is examined, and forestry management makes sure they scour through every neighborhood within the cluste [@StreetTreeMaintenance]r. The Street Tree dataset therefore, is composed of these clusters, and every street tree found within that cluster. The way clusters are determined is through request by citizens [@StreetTreeMaintenance]. This does mean that some clusters may overlap with each other, especially as requests can be made in the exact same location several times. However, trees that are recorded multiple times will stand to be more accurate over time. This is because the dataset is constantly changing. It is not a snapshot of street trees in Toronto at a specific time. It is meant to be an up-to-date representation of street trees in Toronto. Trees that aren't recorded as often may have their records be more out of date. For example, dbh_trunk is a varaible that will slowly increase with time (barring tree removal or extreme modifications otherwise). A record 2 or 3 years behind the current date would display a dbh_trunk value that may not reflect the current value. This may be a reason ASTM is looking towards city-wide periodic checks, with reports suggesting tree pruning every seven years.

`botanical_name` is derived when Forestry Management successfully identifies the species of a tree. The scientific name is recorded in the dataset, along with it's common English name (but this report doesn't use the English name). The location of the tree is then allocated to a postal address of a (private) property. This does not mean the tree is located in this postal address, just that it is in proximity to it. A postal address will have a streetname, recorded as `streetname`, as well as unit number (which isn';'t used in this paper). Given the postal address of a street tree, crosstreets can be determined by map, or by on-site teams overseeing any maintenance at that location. This is how `crossstreet1` and `crossstreet2` are documented. Finally, using a postal address allows for the tree to be located by ward. Boundaries for each ward are listed on the City of Toronto's website, allowing for easy categorization.

@@

Next steps would be augmenting this data with other data from Toronto's roads, or other trees, to see if any comparisons or correlations can be drawn there. Does vehicle traffic impact the trees lining a certain road, and if so, how long would it take for the effects to be noticeable? Does the distribution of street trees tell us anything about income disparity, fire risk, or quality of life within Toronto? This report can be used as a launching pad for further investigations, using key takeaways from this report as drivers for further questions.

With newer report, pressure will be put upon Urban Forestry to better curtail the data they have collected thus far. It is imperative that their hard work be recognized and any feedback on the data they have collected be sent to them as well, ensuring better datasets for analysts, leading to more informative conclusions for all.

@@

## Surveys, sampling, and observational data

ASTM's survey of Toronto's street trees leaves a lot to be desired. With their implementation of cluster-sampling, certain "clusters" of the city may be left unrecorded for extended periods of time. This introduces an element of variance with respect to time. Certain trees have more up-to-date assesments than others. The way this impacts

@@

While cleaning that dataset, some inconsistent values were discovered. A particular species of tree, Salix x sepulcralis, had it's name misspelled in the dataset. This may have come from using special characters that could not be saved onto the dataset, resulting in a garbled name. An additional step was taken to restore the correct botanical name, using the common name as reference. Similarly, one observation did not have a ward location allocated. This was remedied by reading the global coordinates from geometry, a variable in the raw data file that shares the latitude and longitude of the tree individual. Using these coordinates, and the address number, the general location for this tree was determined, along with its ward location. The ward entry for this observation was updated accordingly.
